\textbf{\supplement}\vskip 2ex \mytitle\vskip 2ex

### 2023-01-31 paper-seb-supplementary-material.tex
Fixing git messages

\iftrue
    \PassOptionsToPackage{hyphens}{url}
    \documentclass[10pt,letterpaper]{article}

    \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
    % Use adjustwidth environment to exceed column width (see example table in text)
    \usepackage{changepage}

    % line numbers
    \usepackage[right]{lineno}
    \linenumbers

    \renewcommand\thesubsection{\thesection.\alph{subsection}}
    %\usepackage{hyperref}
    %\usepackage{refcount}
    \usepackage{url}
    \usepackage{doi}
    \usepackage{longtable}
    \usepackage{color}
    \usepackage{amsmath,amssymb}
    \usepackage{fancyhdr}
    \def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
    \def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}

    \input paper-seb-macros.tex
\fi

\makeatletter
      \renewcommand{\thefootnote}{\arabic{footnote}}
      {\let\oldbibcite=\bibcite
      % example: \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
      \newcount\mainNumberOfReferences \mainNumberOfReferences=0
      \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
      \input paper-seb-main.aux 
      \resetlinenumber[\MainPaperLastLine]
      }

\makeatother

@@

% pageref gives us a hyperlink which looks unsightly
\makeatletter
\def\rememberpage{\immediate\write\@auxout{\bslash gdef\bslash SupplementalLastPage{\thepage}}}
\makeatother

@@

% this will be used to save cross references that can be used in the main paper too
\newwrite\infofile
\immediate\openout\infofile=generated-info-for-main.tex

@@

\makeatletter
\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
\immediate\write\infofile{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
% as well as in the current document....
\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
}
\makeatother

@@

\setcounter{section}{0}
\def\n#1{{\sf\bfseries ({#1})}}

@@

\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
\def\changemysection#1{\section{{#1}}\sectioncontents{{#1}}}
\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
\def\mysection#1{\section{#1}\sectioncontents{#1}}
\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}

@@

\noindent\vbox{% The table of contents
\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
\noindent
\input generated-supplementary.toc
}

\newwrite\contentsfile
\immediate\openout\contentsfile=generated-supplementary.toc

\newpage
\mysection{Further issues for Software Engineering Boards (SEBs)}

@@

\begin{enumerate}\raggedright
\item 
Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at software engineering.

\item
When code is taken seriously, concerns may be raised on programmers' contributions to research, intellectual property rights, and co-authoring \cite{vancouver}. Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves. Often software engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}. Often the software engineers will be solving entirely new problems and contributing to the research. How this is handled needs exploring. How software engineers are appropriately credited and cited for their contributions also needs exploring.

@@

\item
SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice. For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of software engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.

@@

\begin{enumerate}\raggedright
\item 
Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about software engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on). SEBs may also provide mentoring and training.

\item
Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software. During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional software engineering.

@@

\item
Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway. It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional software engineers to avoid this problem. 
\end{enumerate}

@@

SEBs address the problems identified at the laboratory end of doing science; they do not address the processes of review, editorial control, and action based on claimed results. As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced. In other words, improving the professionalization of software engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication. SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional software engineering, but this is a topic the present paper has not addressed. It needs doing.

@@

This \supplement\ provides more explanations and justification for following standard software engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.

The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to software engineering in science \citeeg{cosmos}.

@@

\vbox{\noindent
      \input generated-minicontents.tex
      }

\newwrite\minicontentsfile
\immediate\openout\minicontentsfile=generated-minicontents.tex

\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations. On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it. The first step of software engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.

In practice, requirements and expectations change. Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance. Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice software engineering can be adopted at any stage, particularly by using \emph{reverse engineering}. In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented. This specification, carefully reviewed, is then used as the basis for a more rigorous software engineering process that implements a more reliable version of the system.

@@

{The type and use of mathematics used in software engineering is formal methods. Not using formal methods ensures the resulting code is unsafe and unreliable. Of particular relevance to scientific modeling: there must be an explicit use of formal methods to ensure  mathematical models (such as differential equations) are correctly implemented in code (and to understand the any limitations of doing so).}

Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others). Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone. Standard tools cover verification, static analysis of code, version control, documentation, and so on --- this paper explains why some of these activities are essential for reliable programming below.

Crucially, tools are designed to catch common human errors that we are all prone to. Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort.

Many programming languages and programming environments have integrated features that support formal methods. For example, Hoare's triples \cite{hoare} (and formal thinking based on similar ideas) are readily supported by assertions, as either provided explicitly in a programming language or through a simple API\@. In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program. Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of software engineering has advanced and become more specialized and sophisticated over the decades since. Hoare is also a very good writer.

@@

\begin{change}Interestingly, the only paper reviewed that claimed to do any independent testing \cite{onlyPaperWithChecks} failed to include any testing in its data or code repository, so the testing itself --- the essential quality assurance of the code --- is not open to scrutiny (e.g., the code and the ``independent'' code are likely to contain common code, data, and common bugs).\end{change}

\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
Many popular languages are popular because they are easy to use, which is not the same as being reliable to use. The fewer constraints a language imposes, the easier it \emph{seems\/} to be to program in, but the lack of constraints means the language cannot provide the checks stricter languages do. C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers. (This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer. SPARK Ada is one example of a much more appropriate programming language to use. SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.

@@

{If results cannot be reproduced, has anything reliable been contributed? When a modeling paper presents results from a model, it is important to reproduce those results without using the same code. Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}). Reproducing the same results relying on the same codebase tells you little. The more independent reproductions of results the greater the evidence for belief in the implications.}

@@

Most version control systems would, in addition, enable open source methods so the code could be shared --- and reviewed --- by a wider community. Open source is not a panacea, however; it raises many trade-offs. Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration. Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.

@@

For purely scientific purposes, perhaps the most important form of internal is internal documentation: how to understand how and why the code works. This is different from developer documentation, which is how to \emph{use\/} the code in other programs. For example, code for solving a differential equation needs explaining --- what method does it use, what assumptions does it have? In contrast, the developer documentation for differentiation would say things like it solves ordinary differential equations with parameters $e$ for the function $f$ with the independent variable $x$ in the interval $[u,v]$, or whatever, but \emph{how\/} it solves equations is of little interest to the developer who just needs to use it. How code works --- internal documentation --- is essential for the epidemiologist, or more generally any scientist. An example of a simple SIR epidemiological model's internal documentation can be found at \url{http://www.harold.thimbleby.net/sir}

@@

\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs. The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understanding what useful guarantees these basic properties ensure).

Note that reusing code assumes the originators of the code followed good software engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good software engineering practice, it too can be shared and further improved as it gets more exposure. Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods. Indeed, reuse, scrutiny, and consensus are the foundations of good science.

@@

Note that medical device regulation, which has its own standards, is lagging behind professional software engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.

@@

\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in software engineering is continually evolving}{Continuous Professional Development (CPD)}

As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in software engineering is also rapidly evolving. Continuous Professional Development (CPD) is essential.

@@

Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves. {In software engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on software engineering standards in other fields.}

@@

For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example \cite{example-numerical-error} was noticed to use literal numbers at too high a precision for the chosen language, where conformant implementations use IEEE 754 double precision 64-bit floating point. Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.

{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for software engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}

\mysection{Code, data and publication}
\immediate\write\infofile{\bslash newlabel{on-code-data-publication}{{\arabic{section}}{\thepage}{XXX}{YYYY}{ZZZZ}}}

\begin{change}
All computer systems are in principle equivalent to Turing Machines, and Turing Machines make no distinction between program and data. It is possible to define Turing Machines that do separate program code and data, but as soon as a Universal Turing Machine is constructed, its data \emph{is\/} code. Indeed, Universal Turing Machines are a theoretical abstraction of virtual machines, which are used widely in practical computing. Java, for instance, runs in a virtual machine, so any Java program code (and any data it uses) is in fact merely \emph{all\/} data to the Java virtual machine. At another extreme, $\lambda$-calculus is purely program source code, yet $\lambda$-calculus is equivalent to Turing Machine computation. Therefore, even the ``pure'' programs of $\lambda$-calculus also represent data.
\end{change}

>>>>

\iftrue
    \PassOptionsToPackage{hyphens}{url}
    \documentclass[10pt,a4paper]{article}
    \usepackage[text={6.5in,10in},centering]{geometry}    
    \usepackage{url}
    \usepackage{doi}
    \usepackage{longtable}
    \usepackage{color}
    \usepackage{amsmath,amssymb}
    \usepackage{fancyhdr}
    \usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
    \def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
    \def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}    
    \input paper-seb-macros.tex
\fi

\makeatletter
      \renewcommand{\thefootnote}{\arabic{footnote}}
      \let\oldbibcite=\bibcite
      % example: \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
      \newcount\mainNumberOfReferences \mainNumberOfReferences=0
      \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
      \input paper-seb-main.aux 
      %\resetlinenumber[\MainPaperLastLine]
	  \let\bibcite=\oldbibcite
\makeatother

\input generated/page-lengths.tex

@@

% pageref gives us a hyperlink which looks unsightly, so we do it this way...
\makeatletter
\def\rememberpage{\immediate\write\@auxout{\bslash gdef\bslash SupplementalLastPage{\thepage}}}
\makeatother

@@

% this will be used to save cross references that can be used in the main paper too
\newwrite\infofileForMain
\immediate\openout\infofileForMain=generated/info-for-main.tex

%\makeatletter
%% saves name-value pairs in both the supplementary material .aux file and the info for the main file
%\def\remembernamevalue#1#2{%
%	\immediate\write\@auxout{\@backslashchar gdef\@backslashchar #1{#2}}%
%	\immediate\write\infofileForMain{\@backslashchar gdef\@backslashchar #1{#2}}%
%}
%\makeatother

\input generated/info-for-supplement.tex
\newwrite\infofileForSupplement
\immediate\openout\infofileForSupplement=generated/info-for-supplement.tex

@@

\makeatletter
\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
\typeout{Global globalLabel: #1}%
\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
% as well as in the current document....
\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
}
\def\globalTableLabel#1{% we also need journal code-policies to be used in the main paper...
\typeout{Global globalTableLabel: #1}%
\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{table}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
% as well as in the current document....
\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{table}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
}
\def\globalLabelValue#1#2{% we also need journal code-policies to be used in the main paper...
\typeout{Global globalLabelValue: #1 := #2}%
\immediate\write\infofileForMain{\bslash newlabel{#1}{{#2}{\thepage}{AAA}{BBB}{CCC}}}
}
\makeatother

@@

%\setcounter{section}{0}
\setcounter{section}{\continueSectionNumbers}
\def\n#1{{\sf\bfseries ({#1})}}

@@

\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
\def\changemysection#1{\vskip 2ex\section{{#1}}\sectioncontents{{#1}}}
\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
\def\mysection#1{\section{#1}\sectioncontents{#1}}
\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}

@@

\noindent\vbox{% The table of contents
\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
\noindent\small
\input generated/supplementary.toc
}

\newwrite\contentsfile
\immediate\openout\contentsfile=generated/supplementary.toc

\newpage
\begin{center}
\emph{Pagination, section, and citation numbering continues from the main paper}
\end{center}
\mysection{Further issues for Software Engineering Boards (SEBs)}

@@

\begin{enumerate}\raggedright
\item 
Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at Software Engineering.

\item
When code is taken seriously, concerns may be raised on programmers' contributions to research, intellectual property rights, and co-authoring \cite{vancouver}. Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves. Often Software Engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}. Often the Software Engineers will be solving entirely new problems and contributing to the research. How this is handled needs exploring. How Software Engineers are appropriately credited and cited for their contributions also needs exploring.

@@

\item
SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice. For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of Software Engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.

@@

\begin{enumerate}\raggedright
\item 
Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about Software Engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on). SEBs may also provide mentoring and training.

\item
Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software. During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional Software Engineering.

@@

\item
Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway. It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional Software Engineers to avoid this problem. 
\end{enumerate}

@@

SEBs address the problems identified at the laboratory end of doing science; they do not address the processes of review, editorial control, and action based on claimed results. As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced. In other words, improving the professionalization of Software Engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication. SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional Software Engineering, but this is a topic the present paper has not addressed. It needs doing.

@@

This \supplement\ provides more explanations and justification for following standard Software Engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.

The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to Software Engineering in science \citeeg{cosmos}. Written and maintained by a team of experts, a substantial and wide-ranging reference is the Software Engineering Body of Knowledge (SWEBOK) \cite{swebok}, recognized as International Standards Organization Technical Report 19759.

The Turing Institute has an excellent open resource \cite{turing-way}, though it emphasizes RAP for handling data and authoring papers rather than for programming reliably.

@@

\vbox{\noindent
      \input generated/minicontents.tex
      }

\newwrite\minicontentsfile
\immediate\openout\minicontentsfile=generated/minicontents.tex

\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations. On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it. The first step of Software Engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.

In practice, requirements and expectations change. Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance. Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice Software Engineering can be adopted at any stage, particularly by using \emph{reverse engineering}. In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented. This specification, carefully reviewed, is then used as the basis for a more rigorous Software Engineering process that implements a more reliable version of the system.

@@

{The type and use of mathematics used in Software Engineering is called \textbf{formal methods}. Not using formal methods ensures the resulting code is unsafe and unreliable. Of particular relevance to scientific modeling: there must be an explicit use of formal methods to ensure mathematical models (such as differential equations) are correctly implemented in code (and to understand the any limitations of doing so).}

It is important to be clear that formal methods is a spectrum, from doing it as rigorously and comprehensively as possible using state of the art methods, to applying a basic ``formal methods mindset'' that any competent programmer could do.

\begin{itemize}\raggedright\item An interesting paper explores a formal approach to coding differential equations \cite{differential-equations}, but it makes it clear that their approach is beyond most programmers. Such formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK \cite{spark}, and \emph{many\/} others). Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone. Standard tools cover verification, static analysis of code, version control, documentation, and so on --- this paper explains why some of these activities are essential for reliable programming below.

\item At the other end of the spectrum, a formal methods mindset means being clear about basic mathematical properties of code, such as by using assertions, proving invariants in loops, and so on. This approach should be feasible --- and perhaps required --- for all scientific programmers.
\end{itemize}

Crucially, computer tools are available to catch common human errors that we are all prone to. Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort \cite{misra}.

Many programming languages and programming environments have integrated features that support formal methods. For example, Hoare's triples \cite{hoare} (and formal thinking based on similar ideas) are readily supported by assertions, as either provided explicitly in a programming language or through a simple API\@. In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program.

Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of Software Engineering has advanced and become more specialized and sophisticated over the decades since. Hoare is also a very good writer.

@@

Interestingly, the only paper reviewed that claimed to do any independent testing \cite{onlyPaperWithChecks} failed to include any testing in its data or code repository, so the testing itself --- the essential quality assurance of the code --- is not open to scrutiny (e.g., the code and the ``independent'' code are likely to contain common code, data, and common bugs).

\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
Many popular languages are popular because they are easy to use, which is not the same as being reliable to use. The fewer constraints a language imposes, the easier it \emph{seems\/} to be to program in, but the lack of constraints means the language cannot provide the checks stricter languages do. C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers. (This is not the place for a review of bad programming languages, for which see \cite{heedless}, but Excel is even worse than C\@.)

In particular, C is not a portable language, which means C code will work differently on different types of computer and operating system.

SPARK Ada is a popular example of a much more appropriate high integrity programming language to use \cite{spark}. SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.

Other high integrity languages include  OCaml, F*, and Haskell; reference \cite{haskell} is an excellent introduction to Haskell, and introduces the wider issues of reliable programming in such languages.

@@

{If results cannot be reproduced, has anything reliable been contributed? When a modeling paper presents results from a model, it is important to reproduce those results without using the same code. Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using \emph{Mathematica\/} --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}). Reproducing the same results relying on the same codebase tells you little. The more independent reproductions of results the greater the evidence for belief in the implications.}

@@

Most version control systems would, in addition, enable open source methods so the code could be shared --- and reviewed --- by a wider community. Open source is not a panacea, however; it raises many trade-offs. Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration. Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and as there are.are serious dependability problems with Excel (illustrated particularly in section \ref{benefits-beyond-science} of the main paper), open source projects competently implemented (e.g., avoiding or carefully managing the use of Excel) would help these people enormously.

@@

For purely scientific purposes, perhaps the most important form of documentation is internal documentation: how to understand how and why the code works. This is different from developer documentation, which is how to \emph{use\/} the code in other programs. For example, code for solving a differential equation needs explaining --- what method does it use, what assumptions does it have? In contrast, the developer documentation for differentiation would say things like it solves ordinary differential equations with parameters $e$ for the function $f$ with the independent variable $x$ in the interval $[u,v]$, or whatever, but \emph{how\/} it solves equations is of little interest to the developer who just needs to use it. How code works --- internal documentation --- is essential for the epidemiologist, or more generally any scientist. An example of a simple SIR epidemiological model's internal documentation can be found at \url{http://www.harold.thimbleby.net/sir}

@@

\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs. The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understand what useful guarantees these basic properties ensure).

Note that reusing code assumes the originators of the code followed good Software Engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good Software Engineering practice, it too can be shared and further improved as it gets more exposure. Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods. Indeed, reuse, scrutiny, and consensus are the foundations of good science.

@@

Note that medical device regulation, which has its own standards, is lagging behind professional Software Engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.

@@

\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in Software Engineering is continually evolving}{Continuous Professional Development (CPD)}

As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in Software Engineering is also rapidly evolving. Continuous Professional Development (CPD) is essential.

@@

Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves. {In Software Engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on Software Engineering standards in other fields.}

@@

For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example, one of the surveyed paper's code \cite{example-numerical-error} uses literal numbers at far too high a precision for the chosen language to be able to represent correctly (conformant implementations use IEEE 754 double precision 64-bit floating point). Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running. The error belies misunderstandings in programming which may have wider effects, such as consequences of relying on the precision being higher than it is.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.

{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for Software Engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}

\highlightsec{14}{Human Factors are the foundation of everything}{Software is a human activity}

Software is a human activity, and humans are fallible. Even the Software Engineering methodologies to developer better software are themselves human constructs, and are therefore subject to the same fallibilities.

People would generally not make software errors if there were aware they were making errors. Unfortunately programming is a very demanding activity, which causes tunnel vision (also known as loss of situational awareness). Humans have limited cognitive capacity, and programming (especially programming in a competitive environment, like science) drives programmers to use as much of their cognitive skills for the task in hand. The consequence is programmers focus on ``the'' problem as it appears in the code, and inevitably become unaware they are not considering wider issues. The correctness, generality, ethics, and usability of a program are therefore often unintentionally sacrificed to making code work at all.

Confirmation bias is a standard Human Factors problem \cite{fixit}, which encourages us to perform tests that show our programs work. Instead, we should be rigorously testing ways in which programs can fail as well. This is exactly the same issue pointed out by Popper \cite{popper-conjectures-refutations}: scientists should experiment to find reasons why hypotheses are false, and indeed use simple hypotheses that are testable. Software is really no more than a collection of sophisticated hypotheses, and Computer Science is a science of the artificial \cite{simon-artificial}.

Standard Human Factors mitigations for such problems include team working, with appropriate precautions to manage authority gradients (where the Human Factors oversights of the leader influence the team). Many computerized mitigations are also available --- strong typing, code analyzers, formal methods, and so on, as described in this section of the \supplement.

Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic Software Engineering. Dunning and Kruger go on to say,

\begin{quote}\sf\raggedright\setbox0=\hbox{``}\hskip -\wd0\copy0 People usually choose what they think is the most reasonable and optimal option \hbox{[\hskip 2pt\ldots]} The failure to recognize that one has performed poorly will instead leave one to assume that one has performed well; as a result, the incompetent will tend to grossly overestimate their skills and abilities. \hbox{[\hskip 2pt\ldots]} Not only do these people reach erroneous conclusions and make unfortunate choices, but their incompetence robs them of the metacognitive ability to realize it.''\end{quote}

Unlike many skills (skating, brain surgery, \ldots) programming, typical of much engineering, is one where errors can go unnoticed for long periods of time --- things seem to work nicely right up to the moment they fail. The worse programmers are, the more trivial bugs they tend to make, but trivial bugs are easy to find so, ironically, being a poor programmer \emph{increases\/} one's self-assessment because debugging seems very productive. It is easy for poor programmers and their associates to believe they are better than they actually are, fertile ground for the better-than-average bias \cite{dunning-kruger}.

It sounds harsh to call programmers incompetent, but challenged with the complexity of programs and the complexity of the domains programs are applied in, we are all incompetent and succumb to the limitations of our cognitive resources, suffering blindspots in our thinking \cite{fixit}. We \emph{all\/} make mistakes we are unaware of. If we do not have the benefit of professional qualifications that have assessed us objectively, we generally have a higher opinion of our own competence than is justified. Moreover, if we do not work in a diverse team, nobody will ever point this out, so the potential problems it causes will never be addressed.

{Everyone is subject to Human Factors (including the author of the present paper, e.g., as discussed in \cite{enigma}): for instance, the standard cognitive bias of confirmation bias encourages us to look for bugs when code fails to do what is expected and then debug it to produce better results, but if code generates expected results not to bother to debug it further. This of course tends to make code increasingly conform to prior expectations, whether or not those expectations are scientifically justified. Typically, there was no prior specification of the code, so the code should be right, especially after all the debugging to make it ``correct''! Thus coding routinely suffers from HARKing (Hypothesizing After the Results are Known \cite{harking}), a methodological trap widely recognized in statistics.}

Computers themselves are also a part of the problem. Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable. Programs can be written so that it is not possible to determine what they do or how they do it (whether by deliberate obfuscation, as in malware, or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}. The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).

\mysection{Code, data, and publication}
\globalLabel{on-code-data-publication}
